Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2018-10-12 FreeLogic/2018-JAR/article.tex
modfied

\newcommand\map[1]{\hat{#1}}
\newcommand\hol[1]{\boldsymbol{#1}}
\newcommand\lift[1]{\lceil #1 \rceil}
\newcommand\llift[1]{\dot{#1}}

@@

@@

\author{Christoph Benzm\"uller  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   University of Luxembourg, Luxembourg \& Freie Universit\"at Berlin, Berlin, Germany \\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of California, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

\section{Introduction}
\label{intro}
Partiality and undefinedness are prominent challenges in various areas
of mathematics and computer science.  Unfortunately, however, modern
proof assistant systems and automated theorem provers based on
traditional classical or intuitionistic logics provide rather
inadequate support for these challenge concepts.  Free logic \cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} offers a
theoretically appealing solution, but is has been considered as rather
unsuited towards practical utilization.

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free logic in HOL. We present, as an
example, one implementation of this idea in the mathematical proof
assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic (respectively
inclusive logic) 
\cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subseteq \D$ is
chosen to characterize the subdomain of ``\emph{existing}'' resp. ``\emph{defined}''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only.

@@

The particular notion of free logic as exploited in the remainder of
this article has been proposed by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.  
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.

@@

\begin{definition}[Syntax of $\FFOL$]
  We start with a denumerable set $\V$ of variable symbols, a denumerable set
  $\F$ of $n$-ary function symbols ($n\geq 0$), and a denumerable set
  $\P$ of $n$-ary predicate symbols ($n\geq 0$).

  The \emph{terms and formulas of $\FFOL$} are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x\lambdot r$ are (compound) formulas of
  $\FFOL$, and
\item given a formula $r$ of $\FFOL$, then $\Descr x\lambdot r$ is a
  term of $\FFOL$ (definite  description).
\end{enumerate}
\end{definition}

@@

\begin{definition}[Model of $\FFOL$]
  A \emph{model (structure) for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subseteq D$ a possible empty set of existing/defined objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. $\True$ or $\False$ denote truth and
  felshood respectively. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object. 
\end{definition}

\begin{definition}[Evaluation Function for $\FFOL$]
  The \emph{value} $\| s \|^{M,g}$ of a term or formula $s\in\FFOL$
  in a model $M=\langle D, \ex, I, \star\rangle$ under assignment $g$ 
  defined in the following way:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{M,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{M,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{M,g} = I(f)(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x\lambdot r \|^{M,g} = d \in \ex$, such that $\| r
  \|^{M,g[d/x]} = \True$ and  $\| r
  \|^{M,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x\lambdot r \|^{M,g} = \star$
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}

@@

\end{definition}

\begin{definition}[Validity]\label{ffollvalid}
  A formula $s_o$ is \emph{true} in model $M$ under assignment $g$ if and only if
  $\|s_o\|^{M,g} = T$; this is also denoted as $M,g \models^\FFOL s_o$.  A
  formula $s_o$ is called \emph{valid} in $M$, which is denoted as
  $M \models^\FFOL s_o$, if and only if $M,g \models^\FFOL s_o$ for all assignments
  $g$. Finally, a formula $s_o$ is called \emph{valid}, which we
  denote by $\models^\FFOL s_o$, if and only if $s_o$ is valid for all
  $M$. 
% Moreover,
%   we write $\Gamma\models^\FFOL \Delta$ (for sets of formulas $\Gamma$ and
%   $\Delta$) if and only if there is a model $M$ and an assignment $g$ such that
%   $M,g \models^\FFOL s_o$ for all $s_o\in \Gamma$ and $M,g \models^\FFOL t_o$ for
%   at least one $t_o\in \Delta$.
\end{definition}

@@

\begin{definition}[Types]\label{holtypes}
  The set $\hol{T}$ of simple types is freely generated from a set of
  basic types $\{\hol{o}, \hol{i}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans and 
  $\hol{i}$ is the type of individuals. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{definition}

\begin{definition}[Syntax of $\HOL$]\label{holgrammar}
The terms of $\HOL$ with primitive equality and
definite description are defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \S1.4 of Andrews's overview
  article~\cite{AndrewsSEP}. The choice here closely reflects the
  set of primitive connectives as chosen in higher-order automated theorem provers such
  as LEO-II \cite{LEO}, Leo-III \cite{Leo-III} and Satallax \cite{Satallax}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
\emph{formulas}.  The \emph{logical connectives} of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{=_{\alpha\ar\alpha\ar  o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (where $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{definition}

@@

%\begin{definition}\label{homlbetaeta} %\footnote{\textbf{Chris:} Is Def.4 needed at all?}
  % Two common relations on terms are given by $\hol{\beta}$-reduction and
  % $\hol{\eta}$-reduction.  A $\hol{\beta}$-redex has the form $\hol{(\lam{X}s)t}$ and
  % $\hol{\beta}$-reduces to $\hol{[t/X]s}$.  An $\hol{\eta}$-redex has the form
  % $\hol{(\lam{X}s X)}$ where variable $\hol{X}$ is not free in $\hol{s}$; it
  % $\hol{\eta}$-reduces to $\hol{s}$.  We write $\hol{s\eqb t}$ to mean $\hol{s}$ can be
  % converted to $\hol{t}$ by a series of $\hol{\beta}$-reductions and expansions.
  % Similarly, $\hol{s\eqbe t}$ means $\hol{s}$ can be converted to $\hol{t}$ using both
  % $\hol{\beta}$ and $\hol{\eta}$.  For each $\hol{s_\alpha\in \HOL}$ there is a unique
  % \emph{$\hol{\beta}$-normal form} and a unique \emph{$\hol{\beta\eta}$-normal
  %   form}.
% From this fact we know $s\eqb t$ ($s\eqbe t$) iff
% $\Bnormform s\Metaeq\Bnormform t$
% ($\Benormform s\Metaeq\Benormform t$).
% The semantics of $\stt$ is well understood and thoroughly documented
% in the literature \cite{Henkin50,Andrews72b,Andrews72a,BBK04}; our
% summary below is adapted from Andrews~\cite{sep-type-theory-church}.
%\end{definition}

%\begin{definition}\label{homlassignment}
A \emph{variable assignment} $\hol{g}$ maps
variables $\hol{X_\alpha}$ to elements in $\hol{D_\alpha}$. $\hol{g[d/W]}$ denotes the
assignment that is identical to $\hol{g}$, except for variable $\hol{W}$, which is
now mapped to $\hol{d}$.
%\end{definition}

\begin{definition}[Frame for \HOL]\label{homlframe}
A \emph{frame} $\hol{D}$ is a collection $\hol{\{D_\alpha\}_{\alpha\in\entity{T}}}$
of nonempty sets $\hol{D_\alpha}$, such that $\hol{D_o} = \{\True,\False\}$
(where, as before, $\True$ and $\False$ denote truth and falsehood).  The
$\hol{D_{\alpha\ar\beta}}$ are collections of functions mapping
$\hol{D_\alpha}$ into $\hol{D_\beta}$. 
\end{definition}

\begin{definition}[Model for \HOL]\label{holmodel}
  A \emph{model (structure)} for \HOL is a tuple $\hol{M}=\hol{\langle D, I \rangle}$, where
  $\hol{D}$ is a frame, and $\hol{I}$ is a family of typed interpretation
  functions mapping constant symbols $\hol{p_\alpha}$ to appropriate
  elements of $\hol{D_\alpha}$, called the \emph{denotation of $\hol{p_\alpha}$}
  (the logical connectives $\hol{\neg}$, $\hol{\vee}$, and $\hol{\forall}$ are always
  given the standard denotations, see below).  Moreover, we assume that the domains
  $\hol{D_{\alpha\ar\alpha\ar o}}$ contain the respective identity relations.
\end{definition}

>>>>

\newcommand\map[1]{\widehat{#1}}
\newcommand\hol[1]{\boldsymbol{#1}}
\newcommand\lift[1]{\lceil #1 \rceil}
\newcommand\llift[1]{\dot{#1}}

@@

\newtheorem{thm}{Theorem} 
\newtheorem{defn}[thm]{Definition}
\newtheorem{lem}[thm]{Lemma}

@@

\author{Christoph Benzm\"uller  and Dana S. Scott}
\institute{Christoph Benzm\"uller \at
   Freie Universit\"at Berlin, Berlin, Germany \& University of Luxembourg, Luxembourg\\
%   Stanford University, CSLI/Cordura Hall, Stanford, CA 94305-4115, U.S. \\
   \email{c.benzmueller@gmail.com} \\
% \\
 %            \emph{Present address:} of F. Author  %  if needed
Dana S. Scott \at
Visiting Scholar at University of California, Berkeley,  USA \\
  \email{dana.scott@cs.cmu.edu}
}

@@

\section{Introduction}
\label{intro}
Partiality and undefinedness are prominent challenges in various areas
of mathematics and computer science.  Unfortunately, however, modern
proof assistant systems and automated theorem provers based on
traditional classical or intuitionistic logics provide rather
inadequate support for these challenge concepts.  Free logic \cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} offers a
theoretically appealing solution, but it has been considered as rather
unsuited towards practical utilization.

In the first part of this article (\S2 and \S3) we show how free logic
can be elegantly ``implemented'' in any theorem proving system for
classical higher-order logic (HOL) \cite{B5}. The proposed solution
employs a semantic embedding of free logic in HOL. We present, as an
example, one implementation of this idea in the proof
assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners, such as SMT solvers and
first-order and higher-order automated theorem
provers.\footnote{Cf.~\S\ref{sec-remarks-experiments} for further
  information.} As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

@@

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic (respectively
inclusive logic) 
\cite{Lambert60,Scott67,lambert02:_free_logic,sep-logic-free} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an (existing) object
in a given (non-empty) domain $\D$, and that $\D$ is also exactly the set the
quantifiers range over. In free logic these
basic assumptions are abolished. Terms do still denote objects in a
(non-empty) domain $\D$, but a (possibly empty) set $\ex\subset \D$ is
chosen to characterize the subdomain of ``\emph{existing}'' resp. ``\emph{defined}''
objects in $\D$. Quantification is
now restricted to set $\ex$ of existing/defined objects only.

@@

The particular notion of free logic as exploited in the remainder of
this article has been proposed by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.\footnote{The $\star$
  symbol is not to be confused with any other symbol in Isabelle/HOL.}
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.

@@

\begin{defn}[Syntax of $\FFOL$]
  We start with a denumerable set $\V$ of variable symbols, a denumerable set
  $\F$ of $n$-ary function symbols ($n\geq 0$), and a denumerable set
  $\P$ of $n$-ary predicate symbols ($n\geq 0$).

  The \emph{terms and formulas of $\FFOL$} are formally defined as the
  smallest sets such that:
\begin{enumerate}
\item each variable $x\in\V$ is a term of $\FFOL$, 
\item given any $n$-ary ($n\geq 0$) function symbol $f\in\F$ and terms
  $t_1,\ldots,t_n$ of $\FFOL$, then $f(t_1,\ldots,t_n)$ is a term of $\FFOL$, 
\item given terms $t_1$ and $t_2$ of $\FFOL$, then $t_1 = t_2$ is an
  (atomic) formula of $\FFOL$,
\item given any $n$-ary ($n\geq 0$) predicate symbol $p\in\P$ and terms
$t_1,\ldots,t_n$ of $\FFOL$, then $p(t_1,\ldots,t_n)$ is an (atomic)
formula of $\FFOL$, 
\item given formulas $r$ and $s$ of $\FFOL$, then $\neg r$, $r
  \Impl s$ and $\forall x\lambdot r$ are (compound) formulas of
  $\FFOL$, and
\item given a formula $r$ of $\FFOL$, then $\Descr x\lambdot r$ is a
  term of $\FFOL$ (definite  description).
\end{enumerate}
\end{defn}

@@

\begin{defn}[Model of $\FFOL$]
  A \emph{model (structure) for $\FFOL$} consists of a quadruple
  $M = \langle D, \ex, I, \star\rangle$, where $D$ is a non-empty raw domain of
  objects, $\ex\subset D$ a possible empty set of existing/defined objects, and $I$ an interpretation function
  mapping $0$-ary function symbols (constants) to defined objects $d\in \ex$, 
 $0$-ary predicate symbols (propositions) to $\True$ or $\False$, 
  $n$-ary function symbols (for $n\geq 1$) to $n$-ary functions
  $D \times \ldots \times D \longrightarrow D$ and $n$-ary predicate
  symbols (for $n\geq 1$) to $n$-ary relations
  $D \times \ldots \times D$. $\True$ or $\False$ denote truth and
  felshood respectively. Finally, $\star\in D\setminus \ex$ is a designated
  (non-existing/undefined) object. 
\end{defn}

\begin{defn}[Evaluation Function for $\FFOL$]
  The \emph{value} $\| s \|^{M,g}$ of a term or formula $s\in\FFOL$
  in a model $M=\langle D, \ex, I, \star\rangle$ under assignment $g$ 
  defined in the following way:
\noindent
\begin{description}
\item{Terms}
\begin{enumerate}
\item $\| x \|^{M,g} = g(x)$ for variable symbols $x\in\V$
\item $\| c \|^{M,g} = I(c)$, where $c\in\F$ is an $0$-ary function symbol
\item $\| f(t_1,\ldots,t_n)\|^{M,g} = I(f)(\| t_1 \|^{M,g},\ldots,\|
  t_n \|^{M,g})$, where $f\in\F$ is an $n$-ary ($n\geq 1$) function symbol
\item $\| \Descr x\lambdot r \|^{M,g} = d \in \ex$, such that $\| r
  \|^{M,g[d/x]} = \True$ and  $\| r
  \|^{M,g[d'/x]} = \False$ for all $d'\not=d\in \ex$ (i.e. $d$ is the
  unique existing object for which $r$ holds); if there is no
  such $d\in \ex$, then $\| \Descr x\lambdot r \|^{M,g} = \star$
 \newcounter{enumTemp}
 \setcounter{enumTemp}{\theenumi}
\end{enumerate}

@@

\end{defn}

\begin{defn}[Validity]\label{ffollvalid}
  A formula $s_o$ is \emph{true} in model $M$ under assignment $g$ if and only if
  $\|s_o\|^{M,g} = True$; this is also denoted as $M,g \models^\FFOL s_o$.  A
  formula $s_o$ is called \emph{valid} in $M$, which is denoted as
  $M \models^\FFOL s_o$, if and only if $M,g \models^\FFOL s_o$ for all assignments
  $g$. Finally, a formula $s_o$ is called \emph{valid}, which we
  denote by $\models^\FFOL s_o$, if and only if $s_o$ is valid for all
  $M$. 
% Moreover,
%   we write $\Gamma\models^\FFOL \Delta$ (for sets of formulas $\Gamma$ and
%   $\Delta$) if and only if there is a model $M$ and an assignment $g$ such that
%   $M,g \models^\FFOL s_o$ for all $s_o\in \Gamma$ and $M,g \models^\FFOL t_o$ for
%   at least one $t_o\in \Delta$.
\end{defn}

@@

\begin{defn}[Types]\label{holtypes}
  The set $\hol{T}$ of simple types is freely generated from a set of
  basic types $\{\hol{o}, \hol{i}\}$ using the function type
  constructor $\hol{\ar}$. $\hol{o}$ is the type of Booleans and 
  $\hol{i}$ is the type of individuals. We
  may avoid parentheses if the structure of a complex type is clear in context.
\end{defn}

\begin{defn}[Syntax of $\HOL$]\label{holgrammar}
The terms of $\HOL$ with primitive equality and
definite description are defined by the following
grammar:\footnote{It is well known that we could work with a much smaller set
  of logical connectives, see e.g. \S1.4 of Andrews's overview
  article~\cite{AndrewsSEP}. The choice here closely reflects the
  set of primitive connectives as chosen in higher-order automated theorem provers such
  as LEO-II \cite{LEO}, Leo-III \cite{Leo-III}, and Satallax \cite{Satallax}.}
\begin{align*} 
  \hol{s},\hol{t} \quad ::= \quad & \hol{p_\alpha} \mid \hol{X_\alpha} \mid \hol{(\lam{X_\alpha}
  s_\beta)_{\alpha\ar\beta}} \mid \hol{(s_{\alpha\ar\beta}\,
                                    t_\alpha)_\beta} \mid
                                    \hol{s_\alpha = t_\alpha} \mid
                                    \hol{\neg_{o\ar o}\, s_o} \mid\\
  & \hol{(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)} \mid \hol{\forall_{(\alpha\ar
    o)\ar o}(\lam{X_\alpha} s_o)} \mid \hol{\Descr(\lam{X_\alpha} s_o)}
\end{align*}
where $\hol{\alpha},\hol{\beta}\in \hol{T}$. $\hol{p_\alpha}$ denotes typed constants and
$\hol{X_\alpha}$ typed variables (distinct from $\hol{p_\alpha}$).  Complex typed
terms are constructed via abstraction and application. The type of
each term is given as a subscript.  Terms $\hol{s_o}$ of type $\hol{o}$ are called
\emph{formulas}.  The \emph{logical connectives} of choice are
$\hol{\neg_{o\ar o}}$, $\hol{\vee_{o\ar
  o\ar o}}$, $\hol{=_{\alpha\ar\alpha\ar  o}}$, $\hol{\forall_{(\alpha\ar
  o)\ar o}}$ and $\hol{\Descr_{(\alpha\ar
  o)\ar \alpha}}$ (where $\hol{\alpha\in T}$).  Type subscripts may be dropped if
irrelevant or obvious. Similarly, parentheses may be avoided.  Binder
notation $\hol{\all{X_\alpha} s_o}$ and $\hol{\Descr {X_\alpha}\lambdot s_o}$ is used as shorthand for
$\hol{\forall_{(\alpha\ar o)\ar o}
(\lam{X_\alpha} s_o)}$ and $\hol{\Descr(\lam{X_\alpha} s_o)}$, and infix notation $\hol{s \vee t}$ is employed
instead of $\hol{((\vee s)\, t)}$. From the above connectives, other logical
connectives, such as $\hol{\top}$, $\hol{\bot}$, $\hol{\wedge}$,
$\hol{\Impl}$, $\hol{\equiv}$ and $\hol{\exists}$, can be defined in
the usual way. For example, $\hol{\Impl}$ can be defined as $\hol{\lam{X_o}\lam{Y_o}
  \neg X \vee Y}$ and $\hol{\wedge}$ as $\hol{\lam{X_o}\lam{Y_o}
  \neg (\neg X \vee \neg Y)}$.
\end{defn}

@@

%\begin{defn}\label{homlbetaeta} %\footnote{\textbf{Chris:} Is Def.4 needed at all?}
  % Two common relations on terms are given by $\hol{\beta}$-reduction and
  % $\hol{\eta}$-reduction.  A $\hol{\beta}$-redex has the form $\hol{(\lam{X}s)t}$ and
  % $\hol{\beta}$-reduces to $\hol{[t/X]s}$.  An $\hol{\eta}$-redex has the form
  % $\hol{(\lam{X}s X)}$ where variable $\hol{X}$ is not free in $\hol{s}$; it
  % $\hol{\eta}$-reduces to $\hol{s}$.  We write $\hol{s\eqb t}$ to mean $\hol{s}$ can be
  % converted to $\hol{t}$ by a series of $\hol{\beta}$-reductions and expansions.
  % Similarly, $\hol{s\eqbe t}$ means $\hol{s}$ can be converted to $\hol{t}$ using both
  % $\hol{\beta}$ and $\hol{\eta}$.  For each $\hol{s_\alpha\in \HOL}$ there is a unique
  % \emph{$\hol{\beta}$-normal form} and a unique \emph{$\hol{\beta\eta}$-normal
  %   form}.
% From this fact we know $s\eqb t$ ($s\eqbe t$) iff
% $\Bnormform s\Metaeq\Bnormform t$
% ($\Benormform s\Metaeq\Benormform t$).
% The semantics of $\stt$ is well understood and thoroughly documented
% in the literature \cite{Henkin50,Andrews72b,Andrews72a,BBK04}; our
% summary below is adapted from Andrews~\cite{sep-type-theory-church}.
%\end{defn}

%\begin{defn}\label{homlassignment}
A \emph{variable assignment} $\hol{g}$ maps
variables $\hol{X_\alpha}$ to elements in $\hol{D_\alpha}$. $\hol{g[d/W]}$ denotes the
assignment that is identical to $\hol{g}$, except for variable $\hol{W}$, which is
now mapped to $\hol{d}$.
%\end{defn}

\begin{defn}[Frame for \HOL]\label{homlframe}
A \emph{frame} $\hol{D}$ is a collection $\hol{\{D_\alpha\}_{\alpha\in\entity{T}}}$
of nonempty sets $\hol{D_\alpha}$, such that $\hol{D_o} = \{\True,\False\}$
(where, as before, $\True$ and $\False$ denote truth and falsehood).  The
$\hol{D_{\alpha\ar\beta}}$ are collections of functions mapping
$\hol{D_\alpha}$ into $\hol{D_\beta}$. 
\end{defn}

\begin{defn}[Model for \HOL]\label{holmodel}
  A \emph{model (structure)} for \HOL is a tuple $\hol{M}=\hol{\langle D, I \rangle}$, where
  $\hol{D}$ is a frame, and $\hol{I}$ is a family of typed interpretation
  functions mapping constant symbols $\hol{p_\alpha}$ to appropriate
  elements of $\hol{D_\alpha}$, called the \emph{denotation of $\hol{p_\alpha}$}
  (the logical connectives $\hol{\neg}$, $\hol{\vee}$, and $\hol{\forall}$ are always
  given the standard denotations, see below).  Moreover, we assume that the domains
  $\hol{D_{\alpha\ar\alpha\ar o}}$ contain the respective identity relations.
\end{defn}

\begin{defn}[Evaluation Function for \HOL]\label{holvalue}
  The \emph{value} $\hol{\| s_\alpha\|^{M,g}}$ of a \HOL term
  $\hol{s_\alpha}$ on a model $\hol{M}=\hol{\langle D, I \rangle}$ under assignment $\hol{g}$ is an element $\hol{d}\in \hol{D_\alpha}$
  defined in the following way:
\begin{enumerate}
\item $\hol{\|p_\alpha\|^{M,g}} = \hol{I(p_\alpha)}$ 
\item $\hol{\|X_\alpha\|^{M,g}} = \hol{g(X_\alpha)}$
\item $\hol{\|(s_{\alpha\ar\beta}\, t_\alpha)_\beta\|^{M,g}} = \hol{\|s_{\alpha\ar\beta}\|^{M,g}(\|t_\alpha\|^{M,g})}$
\item $\hol{\|(\lam{X_\alpha}s_\beta)_{\alpha\ar\beta}\|^{M,g}} = $
  the function $\hol{f}$ from $\hol{D_\alpha}$ to $\hol{D_\beta}$ such
  that $\hol{f(d)} = \hol{\|s_\beta\|^{M,g[d/X_\alpha]}}$ for all
  $\hol{d}\in \hol{D_\alpha}$
\item $\hol{\|s_\alpha = t_\alpha\|^{M,g}} = {\True}$ if and only if $\hol{\|s_\alpha\|^{M,g}} = \hol{\|t_\alpha\|^{M,g}} $
\item $\hol{\|(\neg_{o\ar o}\, s_o)_o\|^{M,g}} = {\True}$ if and only if $\hol{\|s_o\|^{M,g}} = {\False}$
\item \sloppy $\hol{\|(({\vee_{o\ar o\ar o}}\,s_o)\,t_o)_o\|^{M,g}} =
  {\True}$ if and only if\, $\hol{\|s_o\|^{M,g}} = {\True}$ or $\hol{\|t_o\|^{M,g}}
  = {\True}$
\item \sloppy $\hol{\|(\forall_{(\alpha\ar o)\ar o}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = {\True}$ if and only if for all $\hol{d}\in
  \hol{D_\alpha}$ we have $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\True}$
\item $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{d}$ if there exists a unique $\hol{d}\in \hol{D_{\alpha}}$ such that
  $\hol{\|s_o\|^{M,g[d/X_\alpha]}} = {\True}$, 
otherwise  $\hol{\|(\Descr_{(\alpha\ar o)\ar \alpha}(\lam{X_\alpha}
    s_o))_o\|^{M,g}} = \hol{e}$ for an arbitrary element $\hol{e}\in \hol{D_{\alpha}}$
%\item $\|(\Box_{o\ar o}\, s_o)_o\|^{M,g,w} = T$ iff for all $v \in W$ with $w R v$ we have  $\|s_o\|^{M,g,v}  = T$
\end{enumerate}
\end{defn}

\begin{defn}[Standard and Henkin models]\label{homlhenkinmodel}
 A model $\hol{M}=\hol{\langle D, I \rangle}$ is called a
 \emph{standard model} if and only if for all $\hol{\alpha,\beta}\in \hol{T}$ we have
 $\hol{D_{\alpha\ar\beta}} = \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$. In a \emph{Henkin model} function spaces are not necessarily
 full. Instead it is only required that $\hol{D_{\alpha\ar\beta}}
 \subseteq \{ f \mid f : \hol{D_\alpha} \longrightarrow \hol{D_\beta}\}$ (for all
 $\hol{\alpha,\beta}\in \hol{T}$) and that the valuation function 
 $\hol{\|\cdot\|^{M,g}}$ from above is total (i.e., every term denotes). Any standard model is obviously
 also a Henkin model. 
\end{defn}
